Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-11-14 Articles/2019-ShannonEntropy/Submissions/AJP/AJP-ShannonEntropyIsVariability.tex
Preparing version for AJP

%\documentclass[prb,preprint]{revtex4-1}

\documentclass{article}
\textheight 8.9 in \textwidth 6.5 in \oddsidemargin -10pt \topmargin -30pt

@@

\author{
	Gabriele Carcassi,
	Julian Barbour and
	Christine A. Aidala 
}

@@

\begin{abstract}
TBD
\end{abstract}

@@

The use of Shannon entropy $-\sum p_i \log p_i$ is well established in very different fields of knowledge such as computer science[cite], physics[cite] and ecology[cite]. Within computer science it is a fundamental element in information theory since Shannon developed his mathematical theory of communication. It has been used extensively within statistical mechanics to link the microscopic description of a system to its macroscopic thermodynamic one. It has been used in ecology to characterize biodiversity and calculate species distributions under given constraints. Entropy maximization has been shown to be a powerful tool.

While the techniques are well developed, it is puzzling that this concept can be applied in such different contexts, given that it assumes different features in each one. For example, in information theory the Shannon entropy for a book will change whether we choose to encode words or letters, which means the quantity does not have a unique value: it depends on an arbitrary choice. This is not problematic for an information theoretic quantity, but it is for a physical quantity. In statistical mechanics the $p_i$ represent probabilities, which means the Shannon entropy is a property of an ensemble of physical systems. This leads some to claim that the Shannon entropy is epistemic in nature and represents ignorance. [Add references] In ecology, however, the $p_i$ represent the fraction of individuals belonging to the $i$th species. There is no probabilistic ensemble to which one can associate a degree of ignorance. Finally, entropy in physics is associated to the notion of states and irreversible time evolution, which are not well defined concepts for the other fields. The same expression, then, seems to have different traits and it is not clear why a single concept would apply in all these cases.

To make matters worse, it is fairly typical to describe entropy with notions such as uncertainty[cite], knowledge[cite], lack of knowledge[cite], disorder[cite]. While some of these characterizations are useful in some contexts to give an intuitive picture, they do not always seem to work. A common objection to the characterization as disorder, for example, is that a uniform distribution corresponds to a homogeneous system, which one would expect to be maximally ordered; yet the distribution maximizes the Shannon entropy, which would indicate maximal disorder. Moreover, these notions remain vague in the sense that the formula is not formally derived from those concepts, and therefore it is not clear why that particular indicator would be an appropriate choice. We are left to wonder: is there really some single underlying concept, that, when applied appropriately in the different fields, reduces to the correct notion of entropy? Or is it just a big coincidence?

We propose that there is indeed a characterization of Shannon's formula that is both more general and more appropriate; that is independent of whether it is applied to sources of messages, ensembles of microstates or animal populations. We will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements}. That is, given a distribution of some elements over some quantity, the Shannon entropy represents how varied, how different from each other, the values are. In fact, it is the only continuous and linear index that measures the degree of variability within a distribution, and those requirements lead naturally to the same three conditions that Shannon set for his formula.\cite{Shannon} In other words, the concept of variability within a distribution leads in a formal way to the correct expression. In all cases where the formula is applied, in fact, we are using distributions as objects of our investigation and the variability within that distribution will be of interest. Yet, the physical meaning of those distributions can be widely different: they may represent the fractional distribution of animals according to their species; they may represent probability distributions of different outcomes; they may represent knowledge of certain facts as in Bayesian statistics. Therefore the variability itself, in the context of these different cases, will necessarily acquire a different meaning. In particular, we will see how this perspective naturally unfolds in the contexts of information theory and statistical mechanics.

As we believe this is the right characterization, we will therefore use the expression Shannon ``variability'', rather than ``entropy'', ``information'' or ``diversity index'', as it tells us exactly what the quantity describes. It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Boltzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. We will reserve the word ``entropy'' for the one dealing with physical states and irreversible processes.

This characterization also explains why the Shannon variability has such a broad range of applicability. As nature presents itself in a multitude of forms and combinations, a fundamental task in science is to recognize which objects belong together and study how their different attributes are expressed in different ways or degrees. They may be galaxies, animals or outcomes of the same process; we may study their shape, behavior or numerical quantities. But the idea of grouping them and trying to understand the source of their variation is a common endeavor in science.\footnote{In fact, determining which collection of objects is significant and what attributes are relevant is a significant part of the work.} Measuring and explaining this variability is an essential task. It is this very variety that is the inspiration for and the subject of scientific investigation. A universe without variety, where all objects were the same and remained the same, is barely conceivable and certainly it would leave no space for scientific investigation. In the words of Leibniz, the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.\footnote{MOVE TO BIBLIO Paragraph 58 of Leibniz's \emph{Mondadology}}

As a final note, we want to stress that the goal of this work is not to present some new technical result. The goal is instead to synthesize all details into a coherent picture that gives them context and a precise relationship to each other. This can help, especially the new practitioners, to avoid common misconceptions. Moreover, while some details presented here may be well known in a particular community, we find that they may be new to another. For instance, we will show an example of Huffmann coding to illustrate the meaning of the numerical value associated with the Shannon variability. This is standard practice in computer science courses, yet it is not in statistical mechanics. Therefore restating all the crucial aspects in one place is beneficial to make the connection between them readily available.

%https://www.researchgate.net/publication/259011656_A_tribute_to_Claude_Shannon_1916-2001_and_a_plea_for_more_rigorous_use_of_species_richness_species_diversity_and_the_'Shannon-Wiener'_Index

%https://web.stanford.edu/~hastie/Papers/maxent_explained.pdf

%https://www.mdpi.com/1099-4300/21/8/794/htm

%https://www.sciencedirect.com/science/article/abs/pii/S0169534714001037

%Leibniz
%We recall, though we have been unable to relocate, a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''. [REF]

%We conclude this section with a word on terminology. In mathematics, Cantor defined a set as ``a collection into a whole $M$ of definite and separate objects''. He called the objects \emph{elements}. We will also use the word `elements' to denote the things we consider but subject to the proviso that they have a distinguishing feature (or features) that allows us to say that \emph{they are of the same kind}. Thus, we consider sets of a restricted type. The point is this. To constitute a set in accordance with Cantor's definition, it is only necessary for its elements to be \emph{definite and separate} objects. This leaves total freedom as to how such objects in the universe are to be parcelled up into specific sets. For example, a set might consist of Mars and the word `go'. We think such latitude misses one of the most significant aspects of the universe, namely, that the things within it can be distinguished by attributes that are much more specific than the two required by Cantor. Our field of investigation goes beyond mere number, which washes out all attributes apart from the two, of being definite and separate, that allow us to count. They were the only attributes that Cantor needed; we need more.

@@

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
We note that these are exactly the requirements Shannon put forth for his expression\cite{Shannon}, and that these requirements lead to only one possible option: $H(p_i) = - \sum p_i \log(p_i)$.

The Shannon variability is a clear and important attribute of a given distribution: it quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon variability represents the variability of the elements in a distribution, transcending what the distribution physically signifies.

@@

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary? All these questions are interesting, all these questions are insightful and the answers to all these questions are expressed through just one thing: the Shannon variability.

If the Shannon expression is a measure of variability, how is it connected to information? What is it measuring and in what units?

@@

We have seen that the Shannon variability measures the variability within a distribution but, in the end, it is a numeric quantity and we cannot expect to understand the concept if we do not understand what the number represents and in what units.

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like ``is it an animal?" would be redundant. Others, like ``is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. However, it is clear that there is always a minimum number of questions that must be asked to get to the answer. Given that there are more than two species, a single question cannot be enough. It should however be intuitively clear that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon variability quantifies: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the dsitribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies how much information is needed to transfer a value from a distribution.

@@

For example, suppose the population of pets in a country is as follows:
\begin{itemize}
\item dogs 27\%
\item cats 48\%
\item fish 10\%
\item bird 8\%
\item small mammal 4\%
\item reptile 3\%
\end{itemize}
For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask ``is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask `` is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptile and bird with small mammal to form a 13/14 split. So we can ask ``is it a fish or a reptile?" If the answer is yes, the followup question would be ``is it a fish?" If the answer is no, the followup question would be ``is it a bird?" With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
\item bird 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
\item small mammal 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
\item reptile 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

Now that we understand that the Shannon variability is measured by the number of bits required to identify an element from a distribution, it is instructive to derive the same expression from different considerations. Suppose we have a sequence of $N$ elements, say pets like in the previous example. These are taken from $I$ different cases: dogs, cats, fish and so on. Suppose $N_i$ are the of elements of each type, which means $\sum_{i \in I} N_i = N$. Then, given a particular instance of $N_i$, we have $W(N_i) = \frac{N!}{\prod_{i \in I} N_i!}$ possible ways to realize that sequence, which corresponds to all possible permutations. If all permutations are equally likely, then $\log W (N_i)$ represents the number of bits needed to identify one of the sequences. Assuming we use base $b$ for the logarithm, we take $N$ to be large and, using Stirling's approximation $\ln N! \approx N \ln N - N$, we find:
\begin{equation}
\begin{aligned}
\log W(N_i) &= \log \frac{N!}{\prod_{i \in I} N_i!} \\
&= \log N! - \sum_{i \in I} \log N_i! \\
&\approx \frac{1}{\ln b} \left(N \ln N - N - \sum_{i \in I} N_i \ln N_i + \sum_{i \in I} N_i \right) \\
&= \frac{1}{\ln b} \left(\sum_{i \in I} N_i \ln N - \sum_{i \in I} N_i \ln N_i \right) \\
&= -\sum_{i \in I} N_i \left(\log N_i - \log N \right) \\
&= - N \sum_{i \in I} \frac{N_i}{N} \log \frac{N_i}{N} \\
&= - N \sum_{i \in I} p_i \log p_i \\
&= N H(p_i)
\end{aligned}
\end{equation}

The result should not be surprising: it simply tells us that encoding a sequence of $N$ elements is the same as encoding $N$ elements one at a time.

@@

The Shannon variability of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The variability is the information needed to go from the distribution, which is given, to an individual element. It is really the information gap from the population to an element. That is why some people say the entropy is ``lack of information", which is justified because, in a way, it is the information about the elements that the distribution cannot provide. But, again, this is just missing the point of what is being described. If one is not interested in identifying elements there is no ``missing information".

@@

This should suggest that the success of Shannon variability and its use in both information theory and physics does not warrant a fundamental change of perspective in what constitutes a physical object, as some physicists have claimed. It is true that any physical process can be used to process information, when properly encoded. It is also true that scientific theories, in the end, are models that can capture only the aspects of nature that can be tested experimentally, the information extracted by the experiment, under suitable circumstances. Therefore the claim that information plays an essential role in physical theories has a valid basis.[cite Wheeler, Landauer] But it is also true that that data requires the context to be able to be understood: we need to know what is the subject of our experiment, how to prepare it and how to collect the data. The art of experimental science is neither contained in the mathematical description nor in the data collected. As information, in the information theoretic sense, requires that context to become intelligible, it cannot play a primary role. Therefore the claim that the universe itself is information does not follow.[cite Vedral, Davies]

@@

A standard quick and dirty way to extend distributions from discrete variables to continuous variables is to substitute weights with densities and the sums with integrals. Therefore, instead of having a discrete normalized distribution $\sum_i p_i = 1$, we have a continuous normalized distribution $\int \rho(x)dx=1$. While simply changing $- \sum_i p_i \log p_i$ to $- \int \rho(x) \log \rho(x) dx$ works in most cases, it is easy to get confused because the two expressions have significantly different properties.

@@

This brings up a second problem: what happens if we change unit? If we change the reference, we expect that the entropy will change accordingly. In fact:
\begin{equation}
\begin{aligned}
H(\rho(y)) &= - \int \rho(y) \log \rho(y) dy \\
&= - \int \rho(x) \left|\frac{\partial y}{\partial x}\right| \log \left(\rho(x)\left|\frac{\partial y}{\partial x}\right| \right) \left|\frac{\partial x}{\partial y}\right| dx \\
&= - \int \rho(x)  \log \left(\rho(x)\left|\frac{\partial y}{\partial x}\right| \right)  dx \\
&= - \int \rho(x)  \log \rho(x)  dx - \int \rho(x)  \log \left|\frac{\partial y}{\partial x}\right|  dx \\
&= H(\rho(x)) - \int \rho(x)  \log \left|\frac{\partial y}{\partial x}\right|  dx \\
\end{aligned}
\end{equation}
In general, the Shannon variability over a continuous variable is not invariant under coordinate transformations. Under translations, the Jacobian $\left|\frac{\partial y}{\partial x}\right|$ is unitary, so the Shannon variability does not change. If we stretch or shrink, if we change scale, the Shannon variability changes to measure the variability at the new scale. If the change is non-linear, the variability over different ranges will be counted differently.

@@

In physics, there is one thing that makes the Shannon variability work: phase-space. What happens in phase space is that, under coordinate transformation $\hat{q}^i = \hat{q}^i(q^j)$, the $d\hat{q}^i = \frac{\partial \hat{q}^i}{\partial q^j } dq^j$ vary like vector components while $d\hat{p}_i = \frac{\partial q^j}{\partial \hat{q}^i } dp_j$ vary like covector components. Therefore the areas $d\hat{q}^i d\hat{p}_i = dq^j dp_j$ remain the same. The Jacobian determiants are unitary and the Shannon variability is invariant.

@@

The importance of phase space is often attributed to Liouville's theorem, that states that areas are conserved under Hamiltonian evolution. While this is true, their invariance under coordinate transformations is more important as it is what makes them physically meaningful in the first place. In fact, how could we have conservation over time if we could not compare it in an observer independent way? This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\int \rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression would change for those particular variables in those coordinate systems. This can be, again, a source of confusion.\cite{Dunkel}

@@

As we saw in the previous sections, Shannon variability is a very general concept that applies to any type of distribution, no matter what the problem domain. It does not require us to define states, processes, reversibility and therefore it is conceptually different and independent from the entropy in thermodynamics and statistical mechanics. Yet it can, and is, applied successfully in statistical mechanics even in systems far from equilibrium.[cite] How does this work?

As statistical mechanics aims to describe the collective behavior of a large number of physical systems (i.e. particles), the Shannon variability can be applied to their distribution and will correspond to the entropy of the system. Yet, we have to understand that there are two distinct ways to apply the concept simply because there are two types of distributions one can be interested in. We may consider the state of the whole system at a given time, and be interested in the distribution of the particles over all possible particle states. This will lead to the Boltzman entropy. We may consider a statistical ensemble, which is a large collection of independent copies of the system found in different states, and consider the distribution of these different instances over the states of the whole system. This will lead to the Gibbs or the Von Neumann entropy depending on whether the system is classical or quantum. Both these types of distributions are used and are therefore of interest, so we will look at both and clarify the potential sources of confusion.

@@

As we said before, we treat continuous variables by comparing to a finite unit. We can pick a unit of phase space small enough such that the density $\rho$ is constant over cells of that size. We express $\rho$ as the number of particles within the chosen unit and divide phase space into cells. Now we can calculate all the possible permutations of the particles within the different cells and, in these circumstances, we will find that $\log W = N H(\rho)$. That is, the number of permutations at that level of precision will be equivalent to the number of particles times the variability of the distribution at that level of precision. In other words, the Boltzmann entropy reduces to $N$ times the Shannon variability of the single-particle distribution for a given microstate.

@@

The different entropies in statistical mechanics, then, all have a tight link to the Shannon variability. The Boltzmann entropy corresponds to the variability of the state of a particle within a given microstate, provided that there are a large number of independent and indistinguishable particles. The Gibbs entropy corresponds to the variability of a classical microstate as constrained by the macrostate. The Von Neumann entropy is similar but corresponds to the quantum case. The characterization we gave to the Shannon formula, then, is readily applicable to statistical mechanics in a natural way. We find it more precise than the common characterizations, such as disorder or lack of knowledge, and, once internalized, it leads to less confusion.

>>>>

\documentclass[prb, twocolumn]{revtex4-1}

%\documentclass{article}
%\textheight 8.9 in \textwidth 6.5 in \oddsidemargin -10pt \topmargin -30pt

@@

\author{Gabriele Carcassi}
\author{Julian Barbour}
\author{Christine A. Aidala}

\begin{abstract}
	The Shannon entropy, one of the cornerstones of information theory, is widely used in physics, particularly in statistical mechanics. Yet its characterization and connection to physics remains vague, leaving ample room for misconceptions and misunderstanding. In this paper we will show that the Shannon entropy can be fully understood as measuring the variability of the elements within a distribution: it characterizes how much variation can be found within a collection of objects. We will see that it is the only indicator that is continuous and linear, that quantifies the number of yes/no questions (i.e. bits) that are needed to identify an element within the distribution, and we will see how applying this concept to statistical mechanics in different ways lead to the Boltzman, Gibbs and Von Neumann entropies.
\end{abstract}

@@

@@

The Shannon entropy $-\sum p_i \log p_i$ is well established in such disparate fields as computer science[cite], physics[cite] and ecology[cite]. It is most often introduced in physics through statistical mechanics, and associated to vague notions, such as uncertainty[cite], knowledge[cite], lack of knowledge[cite] and disorder[cite]. These prevent a proper understanding of the concept and often lead to confusion and misunderstandings.

This paper aims to give a crisp characterization of the Shannon entropy that is intuitive and precise. The general idea is that it measures the variability of the elements within a distribution, which is a general concept applicable to many branches of science and it is independent of the notion of entropy of thermodynamics and statistical mechanics. Therefore we will use the term \emph{``Shannon variability''}, leaving ``entropy'' to the physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability. It measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. We will study how the expression works over continuous variables, and show how phase-space is special as it leaves the Shannon variability invariant under change of coordinates. We will turn to statistical mechanics and apply Shannon variability in different ways to recover connections to the Boltzmann, Gibbs and Von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch over a few of the common misconceptions.

@@

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
These are the same requirements Shannon put forth for his expression\cite{Shannon}, from which he showed to be the only possible choice: $H(p_i) = - \sum p_i \log(p_i)$. This expression quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on.

@@

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary?

If the Shannon expression is a measure of variability, why is it connected to information? How is variability quantified and in what units?

@@

To understand what the numerical value represents, consider this example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose I pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

Not all questions will be able to extract the same amount of information. Some questions, like ``is it an animal?" would be redundant. Others, like ``is it an American Flamingo (Phoenicopterus ruber)?", would give us a lot of information in the positive case but little in the negative case. However, there has to be a minimum number of questions that must be asked to get to the answer. A single question, for example, cannot be enough given that there are more than two species. It should be intuitively clear that to a greater variability within the distribution will correspond a greater number of questions you must ask. That is exactly what the Shannon variability quantifies: the minimum average number of questions one has to ask to identify a value in the distribution. It gives us the number of questions for an ideal strategy for our game of Twenty Questions.

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the dsitribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies how much information is needed to transfer a value picked from a distribution.

@@

For example, suppose the population of pets in a country is as follows:
\begin{table}[h]
\begin{tabular}{lr}
dogs & 27\% \\
cats & 48\% \\
fish & 10\% \\
bird & 8\% \\
small mammal & 4\% \\
reptile & 3\%
\end{tabular}
\end{table}

For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask ``is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask `` is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptile and bird with small mammal to form a 13/14 split. So we can ask ``is it a fish or a reptile?" If the answer is yes, the followup question would be ``is it a fish?" If the answer is no, the followup question would be ``is it a bird?" With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
\item bird 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
\item small mammal 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
\item reptile 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

Now that we understand that the Shannon variability is measured by the number of bits required to identify an element from a distribution, it is instructive to derive the same expression from different considerations. Suppose we have a sequence of $N$ elements, say pets like in the previous example. These are taken from $I$ different cases: dogs, cats, fish and so on. Suppose $N_i$ are the of elements of each type, which means $\sum_{i \in I} N_i = N$. Then, given a particular instance of $N_i$, we have $W(N_i) = \frac{N!}{\prod_{i \in I} N_i!}$ possible ways to realize that sequence, which corresponds to all possible permutations. If all permutations are equally likely, then $\log W (N_i)$ represents the number of bits needed to identify one of the sequences.

When $N$ is large, we can use Stirling's approximation $\ln N! \approx N \ln N - N$, and find $\log W(N_i) \approx - N \sum_{i \in I} \frac{N_i}{N} \log \frac{N_i}{N} = N H(p_i)$. The logarithm of the permutations is $N$ times the Shannon variability. The result should not be surprising: it simply tells us that encoding a sequence of $N$ elements is the same as encoding $N$ elements one at a time.

@@

The Shannon variability of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The variability is quantified by the information needed to go from the distribution, which is given, to an individual element. It is really the information gap from the population to an element. That is why some people say the entropy is ``lack of information", which is justified because, in a way, it is the information about the elements that the distribution cannot provide. But, again, this is just missing the point of what is being described. If one is not interested in identifying elements there is no ``missing information".

@@

The use of information in physics, then, does not warrant a fundamental change of perspective in what constitutes a physical object, as some physicists have claimed. It is true that any physical process can be used to process information, \emph{when properly encoded}. It is also true that scientific theories, in the end, are models that can capture only the aspects of nature that can be tested experimentally, the information extracted by the experiment, under suitable circumstances. Therefore the claim that information plays an essential role in physical theories has a valid basis.\footnote{Comments like ``It is wrong, moreover, to regard this or that  physical quantity as sitting \emph{out there} with this or that numerical value it is absorbed'', "the information thus solicited [by the experiment] make physics and comes in bits" by Wheeler\cite{Wheeler} or ``I am proposing that the ultimate form of the implementable laws of physics requires only operations avaialble (in principle) in our actual universe'' by Landauer\cite{Landauer} go in this direction} But it is also true that that data requires the context to be able to be understood: we need to know what is the subject of our experiment, how to prepare it and how to collect the data. The art of experimental science is neither contained in the mathematical description nor in the data collected. As information, in the information theoretic sense, requires that context to become intelligible, it cannot play a primary role. Therefore the claim that the universe itself is information does not follow.\footnote{Comments like ``[Information is] the fundamental building block of the Universe'' by Vedral or ``[Information] occupies the ontological basement'' by Davies go in this direction.}

@@

A standard quick and dirty way to extend distributions from discrete variables to continuous variables is to substitute weights with densities and the sums with integrals. Therefore, instead of having a discrete normalized distribution $\sum_i p_i = 1$, we have a continuous normalized distribution $\int \rho(x)dx=1$. While simply changing $- \sum_i p_i \log p_i$ to $- \int \rho(x) \log \rho(x) dx$ works in most cases, leaving it at that misses a crucial point: the two expressions have significantly different properties.

@@

This brings up a second problem: what happens if we change unit? If we change the reference, we expect that the entropy will change accordingly. In fact we find
\begin{equation}
\begin{aligned}
H(\rho(y)) &= H(\rho(x)) - \int \rho(x)  \log \left|\frac{\partial y}{\partial x}\right|  dx 
\end{aligned}
\end{equation}
In general, the Shannon variability over a continuous variable is not invariant under coordinate transformations. Under translations, the Jacobian $\left|\frac{\partial y}{\partial x}\right|$ is unitary, so the Shannon variability does not change. If we stretch or shrink, if we change scale, the Shannon variability changes to measure the variability at the new scale. If the change is non-linear, the variability over different ranges will be counted differently.

@@

In physics, what makes the Shannon variability work is phase-space. What happens is that, under coordinate transformation $\hat{q}^i = \hat{q}^i(q^j)$, the $d\hat{q}^i = \frac{\partial \hat{q}^i}{\partial q^j } dq^j$ vary like vector components while $d\hat{p}_i = \frac{\partial q^j}{\partial \hat{q}^i } dp_j$ vary like covector components. Therefore the areas $d\hat{q}^i d\hat{p}_i = dq^j dp_j$ remain the same. The Jacobian determiants are unitary and the Shannon variability is invariant.

@@

The importance of phase space is often attributed to Liouville's theorem, that states that areas are conserved under Hamiltonian evolution. While this is true, their invariance under coordinate transformations is more important as it is what makes them physically meaningful in the first place. This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\int \rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression would change for those particular variables in those coordinate systems. This can be, again, a source of confusion.\cite{Dunkel}

@@

Statistical mechanics aims to describe the collective behavior of a large number of physical systems (i.e. particles), therefore asking what variability is expressed by such a collection is a well posed question. Yet, we have to understand that there are two distinct ways to apply the concept simply because there are two types of distributions one can be interested in. We may consider the state of the whole system at a given time, and be interested in the distribution of the particles over all possible particle states. This will lead to the Boltzman entropy. We may consider a statistical ensemble, which is a large collection of independent copies of the system found in different states, and consider the distribution of these different instances over the states of the whole system. This will lead to the Gibbs or the Von Neumann entropy depending on whether the system is classical or quantum. Both these types of distributions are used and are therefore of interest, so we will look at both.

@@

As we said before, we treat continuous variables by comparing to a finite unit. We can pick a unit of phase space small enough such that the density $\rho$ can be considered constant over cells of that size. We express $\rho$ as the number of particles within the chosen unit and divide phase space into cells. Now we can calculate all the possible permutations of the particles within the different cells and, in these circumstances, we will find that $\log W = N H(\rho)$. That is, the number of permutations at that level of precision will be equivalent to the number of particles times the variability of the distribution at that level of precision. In other words, the Boltzmann entropy reduces to $N$ times the Shannon variability of the single-particle distribution for a given microstate.

@@

The different entropies in statistical mechanics, then, all have a tight link to the Shannon variability. The Boltzmann entropy corresponds to the variability of the state of a particle within a given microstate, provided that there are a large fixed number of independently distributed and indistinguishable particles. The Gibbs entropy corresponds to the variability of a classical microstate as constrained by the macrostate. The Von Neumann entropy is similar but corresponds to the quantum case. The characterization we gave to the Shannon formula, then, is readily applicable to statistical mechanics in a natural way.
